Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/splitting.tex
Move 4.0 over to the new notation.

\textbf{Correctness condition for \bbtwo\ term splitting.}
Intuitively, as suggested in \ref{sec:example}, splitting a
\lang\ program (a \bbtwo\ term) should yield two monostage terms where
evaluating the second term (corresponding to second stage
computations) on the output of the first (first stage computations),
yields the same result as direct evaluation of the \lang\ program.
More precisely, for any $\coltwo e A$ which reduces to a residual $q$
(via $\diatwo e q$), $e$ splits into a syntactically separated program
$\mathcal{P}$ (via the new relation $e \splittwosym \mathcal{P}$)
which is reduced to a residual $q'$ (via the new relation $\mathcal{P}
\sepredtwosym q'$), which is the same as $q$ up to an equivalence.  By
``syntactically separated'' we mean that $\mathcal{P}$ has the form
$\pipeS p l r$, where $p$ (called the {\em precomputation}) is a
monostage term representing all the first-stage computations in $e$,
and $l.r$ (called the {\em resumer}) is a monostage term representing
all the second-stage computations in $e$.  $\sepredtwosym$ is defined
by a single rule, which evaluates the precomputation and substitutes
the resulting value in for $l$ in the resumer:
\[
\infer{\sepredtwo {\pipeS p l r} {[b/l]r}} {\reduce p b}
\]
We call $b$ the {\em boundary value} resulting from the split since it represents the data communicated between the first and second stage computations.
A summary of these operations and the relationship between them is provided on the right side of \ref{fig:splittingSummary}.

\textbf{Correctness condition for \bbonem\ term splitting.} Defining
the correctness requirements for splitting \bbonem\ terms is
challenging because unlike \bbtwo\ terms (which reduce to monostage
residuals that can be directly compared to the monostage result of
$\sepredtwosym$), \bbonem\ terms reduce to partial values and residual
tables that represent computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
%For example, the \bbonem\ term:
%\begin{lstlisting}
%(next {1+2}, grnd{3+4})
%\end{lstlisting}
%reduces (via $\redonesym$) to the residual table and partial value:
%\begin{lstlisting}
%[yhat |-> 1+2] (next {yhat}, grnd{7})
%\end{lstlisting}

@@

We define the requirements for splitting \bbonem\ terms using the
results of masking. Specifically, for any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
$\mval i q$ (via $\vsplito$).  Then $e$ splits into a syntactically
seperated program $\mathcal{P}$ (via the new relation $e \splitonesym
\mathcal{P}$) which is reduced to a masked value $\mval i {q'}$ (via
the new relation $\sepredtwosym$) where $q'$ is the same as $q$ up to
an equivalence.

Similar to \bbtwo\ term splitting, $\mathcal{P}$ takes the form
$\pipeM c l r$ where $c$ (called the {\em combined term}) is a
monostage term encoding all the first-stage computations in $e$, and
$l.r$ (again called the {\em resumer}) is a monostage term encoding
all the second-stage computations in $e$.  $\sepredonesym$ is defined
by a single rule that yields masked values:
\[
\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
\]

Note that since $\coltwo e A$ reduces to an entirely second-stage
residual, its first-stage computations only exist to generate input
for second stage computations.  In contrast, since $\colmix e A$
reduces to a multi-stage residual table and partial value, its first
stage computations serve to both generate the value used as input for
the second stage {\em and also} produce a first-stage value.
Correspondingly, in the $\sepredtwosym$ rule, the precomputation $p$
generated via splitting reduces only to a boundary value $b$ which is
passed to the resumer; but in the $\sepredonesym$ rule, the combined
term $c$ reduces to a tuple containing both the first stage result $i$
and the boundary value $b$. A summary of the relationship between the
results of directly evaluating \lang\ and the results of splitting is
provided at left in \ref{fig:splittingSummary}.

@@

The masking operation converts a partial value into a masked value $\mval i q$,
where $i$ is a monostage value encoding all of the first-stage information of the original partial value,
and $q$ is a monostage term encoding all of the second-stage information of the original partial value.
Remember that masking will not play a role in the final splitting algorithm, 
but instead exists to specify the contract to which splitting at \bbonem\ must adhere.
Accordingly, masking should be thought of as a definition, and as such it is in some parts arbitrary.

>>>>

\textbf{Correctness condition for \bbtwo\ term splitting.}  As
suggested informally in \ref{sec:example}, splitting a \lang\ program
(a \bbtwo\ term) should yield two monostage terms where evaluating the
second term (corresponding to second stage computations) on the output
of the first term (first stage computations), yields the same result
as direct evaluation of the \lang\ program.  More precisely, for any
$\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$), $e$
splits (via the new relation~$\splittwosym$) into a syntactically
separated program $\pipeS p l r$ where $\reduce p b$, and $[b/l]r$ is
the same as $q$ up to an equivalence. The term $p$ generated by the
split (called the {\em precomputation}) is a monostage term
representing all the first-stage computations in $e$, and $l.r$
(called the {\em resumer}) is a monostage term representing all the
second-stage computations in $e$.  We refer to the result of the
precomputation ($b$) the {\em boundary value} of the split since it
represents the data communicated between the first and second stage
computations.  The mechanics of splitting \bbtwo\ terms and their
relationship to the outputs of evaluation is provided at right in
\ref{fig:splittingSummary}.

% $\sepredtwosym$ is defined
% by a single rule, which evaluates the precomputation and substitutes
% the resulting value in for $l$ in the resumer:
%\[
%\infer{{[b/l]r}\equiv q} {\reduce p b}
%\]

\textbf{Correctness condition for \bbonem\ term splitting.} Defining
the correctness requirements for splitting \bbonem\ terms is
challenging because unlike \bbtwo\ terms (which reduce to monostage
residuals that can be directly compared to the resumer),
\bbonem\ terms reduce to partial values and residual tables that
represent computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
%For example, the \bbonem\ term:
%\begin{lstlisting}
%(next {1+2}, grnd{3+4})
%\end{lstlisting}
%reduces (via $\redonesym$) to the residual table and partial value:
%\begin{lstlisting}
%[yhat |-> 1+2] (next {yhat}, grnd{7})
%\end{lstlisting}

@@

We define the requirements for splitting \bbonem\ terms using the
results of masking. Specifically, for any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
$\mval i q$ (via~$\vsplito$).  Then $e$ splits (via the new
relation~$\splitonesym$) into a syntactically separated program
$\pipeM c l r$ where $\reduce c (i,b)$, and $[b/l]r$ is the same as
$q$ up to an equivalence.  The term $c$ generated by splitting (called
the {\em combined term}) is a monostage term encoding all the
first-stage computations in $e$, and $l.r$ (again called the {\em
  resumer}) is a monostage term encoding all the second-stage
computations in $e$.  A summary of splitting \bbonem\ terms is
provided at left in \ref{fig:splittingSummary}.

%$\sepredonesym$ is defined by a single rule that yields masked values:
%\[
%\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
%\]

Note that since $\coltwo e A$ reduces to an entirely second-stage
residual, its first-stage computations only exist to generate input
for second stage computations.  In contrast, since $\colmix e A$
reduces to a multi-stage residual table and partial value, its first
stage computations serve to both generate the value used as input for
the second stage {\em and also} produce a first-stage value.
Correspondingly, the precomputation $p$ generated when splitting
\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
the resumer; but the combined term $c$ resulting from \bbtwo\ term
splitting reduces to a tuple containing both the first stage result
$i$ and the boundary value $b$ (motivating the name {\em combined
term}).

@@

Masking converts a partial value and associated residual table into a masked value $\mval i q$,
where $i$ is a monostage value encoding all of the first-stage information of the original partial value,
and $q$ is a monostage term encoding all of the second-stage information of the original partial value.
Remember that masking will not play a role in the final splitting algorithm, 
but instead exists to specify the contract to which splitting at \bbonem\ must adhere.
Accordingly, masking should be thought of as a definition, and as such it is in some parts arbitrary.
